---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r , echo=TRUE}
data <- read.csv(unz(activity.zip, "activity.csv"))
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
```


## What is mean total number of steps taken per day?
The missing values in the dataset are ignored using the command na.rm=TRUE.\
Total number of steps taken per day:\
```{r , echo=TRUE}
total_steps <- data%>% group_by(date) %>% summarise(total_steps = sum(steps, na.rm = TRUE))
```
Histogram of the total number of steps taken each day:\
```{r , fig.margin=TRUE}
hist(total_steps$total_steps, xlab = "Total steps", main = "Total steps per day" )
```
<br>
Mean and median of the total number of steps taken per day:\
```{r , echo=TRUE}
mean_steps <- data%>% group_by(date) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))
median_steps <- data%>% group_by(date) %>% summarise(median_steps = median(steps, na.rm = TRUE))
mean_steps
median_steps
```
<br>
Mean and median of the total number of steps:\
```{r , echo=TRUE}
mean(mean_steps$mean_steps, na.rm = TRUE)
median(median_steps$median_steps, na.rm = TRUE)
```
## What is the average daily activity pattern?
Calculation of the average number of steps per 5-minute interval:\
```{r , echo=TRUE}
interval_avg <- data%>% group_by(interval) %>% summarise(avg_steps = mean(steps, na.rm = TRUE))
```
Time series plot of the 5-minute interval (x-axis) and the average number of steps averaged across all days (y-axis):\
```{r }
plot(interval_avg$interval, interval_avg$avg_steps, type = "l", xlab = "Time interval (5 min)", ylab = "Average steps", main ="Average steps per interval")
```
<br> Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?:\
```{r , echo=TRUE}
max_steps_interval <- interval_avg[which.max(interval_avg$avg_steps), ]
max_steps_interval$interval
```


## Imputing missing values
Total number of missing values in the dataset, and percentage of missing values:\
```{r , echo=TRUE}
number_Nas <- sum(is.na(data$steps))
percentage_Nas <- mean(is.na(data$steps))
```
The strategy for filling in all of the missing values in the dataset consists in using the mean for that 5-minute interval:\
```{r , echo=TRUE}
new_data <- merge(data, interval_avg, by = "interval", x.all = TRUE)
new_data$steps[is.na(new_data$steps)] <- new_data$avg_steps[is.na(new_data$steps)]
new_df <- new_data %>% select(-avg_steps) %>% arrange(date, interval)
```
New dataset that is equal to the original dataset but with the missing data filled in:\
```{r , echo=TRUE}
total_steps_new <- new_df%>% group_by(date) %>% summarise(total_steps = sum(steps, na.rm = TRUE))
```
Histogram of the total number of steps taken each day:\
```{r}
hist(total_steps_new$total_steps, xlab = "Total steps", main = "Total steps per day" )
```
<br> Mean and median total number of steps taken per day:\
```{r , echo=TRUE}
mean_steps_new <- new_df%>% group_by(date) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))
median_steps_new <- new_df%>% group_by(date) %>% summarise(median_steps = median(steps, na.rm = TRUE))
mean_steps_new
median_steps_new
```
The results are not very different from the results obtained excluding the missing data, because these constitute onle the 13% of the dataset. Mean and median of the total number of steps:\
```{r , echo=TRUE}
mean(mean_steps_new$mean_steps, na.rm = TRUE)
median(median_steps_new$median_steps, na.rm = TRUE)
```



## Are there differences in activity patterns between weekdays and weekends?
The dataset used is the one with the filled missing values.\
New factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.\
```{r , echo=TRUE}
weekend <- c("Saturday", "Sunday")
new_df$days <- factor((weekdays(as.Date(new_df$date)) %in% weekend), labels = c("weekday", "weekend"))
interval_avg_new <- new_df%>% group_by(interval, days) %>% summarise(avg_steps = mean(steps, na.rm = TRUE))
```
Panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days:\
```{r}
plot_new_data <- ggplot(interval_avg_new, aes(interval_avg_new$interval, interval_avg_new$avg_steps)) +
  geom_line() +
  ggtitle("Average steps per interval") +
  xlab("Time interval (5 min)") + ylab("Average steps")
my_plot <- plot_new_data + facet_grid(. ~ days)+ theme_bw()+theme(text = element_text(size=12),panel.grid.major = element_blank(),panel.grid.minor =element_blank(),strip.background = element_blank(),plot.title = element_text(hjust=0.5))+  font("title", size = 14, face = "bold")
print(my_plot)
```


